Reproducible Research: Peer Assessment 2
==========================================
*Divya Shree H P*

*December 28, 2019*

## Impact of Severe Weather Events on Public Health and Economy in the United States

### Synonpsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

### Data
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. File can be downloaded from the course web site:

* Storm Data [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service Storm Data Documentation [https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf]

* National Climatic Data Center Storm Events FAQ [https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf]


### Basic settings
```{r global_options}
knitr::opts_chunk$set(fig.path='Plots/')
```
```{r results='hide',message=FALSE, warning=FALSE}
echo = TRUE  # Always make code visible
options(scipen = 1)  # Turn off scientific notations for numbers
library(R.utils)
library(ggplot2)
library(plyr)
library(gridExtra)
options(warn=-1)
```

### Data Processing
First, we download the data file and unzip it.
```{r}
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "stormData.csv.bz2")
bunzip2("stormData.csv.bz2", overwrite=T, remove=F)
```
Then, we read the generated csv file.
```{r}
stormData <- read.csv("stormData.csv", sep = ",")
dim(stormData)
head(stormData,2)
```
There are 902297 rows and 37 columns in total.
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

```{r}
if (dim(stormData)[2] == 37) {
stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
}
hist(stormData$year, breaks = 30,xlab = "Year",main = "Histogram" )
```  

Based on the above histogram, we see that the number of events tracked starts to significantly increase around 1995. So, we use the subset of the data from 1990 to 2011 to get most out of good records.
```{r}
storm <- stormData[stormData$year >= 1995, ]
dim(storm)
```
Now, there are 681500 rows and 38 columns in total.

#### Impact on Public Health
In this section, we check the number of **fatalities** and **injuries** that are caused by the severe weather events. We would like to get the first 15 most severe types of weather events.
```{r}
sortHelper <- function(fieldName, top = 15, dataset = stormData) {
    index <- which(colnames(dataset) == fieldName)
    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}
fatalities <- sortHelper("FATALITIES", dataset = storm)
injuries <- sortHelper("INJURIES", dataset = storm)
```

#### Impact on Economy
We will convert the **property damage** and **crop damage** data into comparable numerical forms according to the meaning of units described in the code book ([Storm Events](http://ire.org/nicar/database-library/databases/storm-events/)). Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r}
convertHelper <- function(dataset = storm, fieldName, newFieldName) {
    totalLen <- dim(dataset)[2]
    index <- which(colnames(dataset) == fieldName)
    dataset[, index] <- as.character(dataset[, index])
    logic <- !is.na(toupper(dataset[, index]))
    dataset[logic & toupper(dataset[, index]) == "B", index] <- "9"
    dataset[logic & toupper(dataset[, index]) == "M", index] <- "6"
    dataset[logic & toupper(dataset[, index]) == "K", index] <- "3"
    dataset[logic & toupper(dataset[, index]) == "H", index] <- "2"
    dataset[logic & toupper(dataset[, index]) == "", index] <- "0"
    dataset[, index] <- as.numeric(dataset[, index])
    dataset[is.na(dataset[, index]), index] <- 0
    dataset <- cbind(dataset, dataset[, index - 1] * 10^dataset[, index])
    names(dataset)[totalLen + 1] <- newFieldName
    return(dataset)
}
storm <- convertHelper(storm, "PROPDMGEXP", "propertyDamage")
storm <- convertHelper(storm, "CROPDMGEXP", "cropDamage")
names(storm)
options(scipen=999)
property <- sortHelper("propertyDamage", dataset = storm)
crop <- sortHelper("cropDamage", dataset = storm)
```

### Results
As for the impact on public health, we have got two sorted lists of severe weather events below by the number of people badly affected.
```{r}
fatalities
injuries
```
We graph the top 10 causes of fatalities. 
```{r}
ggplot(data=fatalities[1:10,], aes(x=EVTYPE, y=FATALITIES)) + 
    geom_bar(stat="identity", fill = "#CC79A7", colour = "Black") + xlab("Event type") + ylab("Total fatalities") + ggtitle("Fatalities By Event Type") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```  

We do the same for injuries.
```{r}
ggplot(data=injuries[1:10,], aes(x=EVTYPE, y=INJURIES)) + 
    geom_bar(stat="identity", fill = "#E69F00", colour = "Black") + xlab("Event type") + ylab("Total injuries") + ggtitle("Injuries By Event Type") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```  

Based on the above plots, we find that **excessive heat** and **tornado** cause most fatalities; **tornato** causes most injuries in the United States from 1995 to 2011.

As for the impact on economy, we have got two sorted lists below by the amount of money cost by damages.  

```{r}
property
crop
```

And the following is a pair of graphs of total property damage and total crop damage affected by these severe weather events. 
```{r}
ggplot(data=property[1:10,], aes(x=EVTYPE, y=propertyDamage)) + 
    geom_bar(stat="identity", fill = "#009E73", colour = "Black") + xlab("Severe Weather Type") + 
    ylab("Property Damage in US dollars") +  ggtitle("Total Property Damage by Severe Weather Events") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```  
```{r}
ggplot(data=crop[1:10,], aes(x=EVTYPE, y=cropDamage)) + 
    geom_bar(stat="identity", fill = "#CC79A7", colour = "Black") + xlab("Severe Weather Type") + 
    ylab("Crop Damage in US dollars") +  ggtitle("Total Crop Damage by Severe Weather Events") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
``` 

Based on the above plots, we find that **flood** and **hurricane/typhoon** cause most property damage; **drought** and **flood** causes most crop damage in the United States from 1995 to 2011.